Clique and Constraint Models for Maximum Common (Connected) Subgraph Problems

### 2016-04-07 paper/paper.tex
Terminology

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two approaches using modern algorithms, and see that the best
    choice depends mainly upon whether the graphs have labelled edges. We also investigate a
    restricted branching approach for the connected version of the problem, show how to implement
    this in constraint-based and clique-inspired algorithms, and compare it to conventional
    propagation.
\end{abstract}

@@

Another widely studied approach ?? cite is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph \cite{LeviG}; we describe
this below. Conventional wisdom is that this approach is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph approach using a modern
algorithm.

@@

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05} to the model. Alternatively, we
could use a special branching rule to maintain connectedness during search ?? cite. Neither approach
seems directly viable with an association graph encoding---however, we show that it is possible to
adapt the combined branching and bounding rule used by modern clique algorithms to preserve
connectedness during search.

Our experiments suggest that each approach gives broadly similar performance, although the
association graph works better when graphs have edge labels (and we can explain why), and the
constraint programming approach is better when edges are unlabelled. For the constraint programming
model, using both a global constraint \emph{and} the branching rule in conjunction seems to be the
safest option---although it is not always fastest, it is very rarely substantially worse than just
using one or the other, and is often substantially better.

@@

    \centering
    \input{gen-graph-unconnected-heatmap}
    \caption{The cumulative number of instances solved in under a certain time: on the top, 33\%
        labelled graphs with up to 100 vertices, and in the middle, unlabelled and undirected graphs
        with up to 35 vertices. On the bottom row, an instance-by-instance comparison of the two
        approaches, with 33\% labelled graphs on the left, and unlabelled and undirected graphs on
        the right.} \label{figure:unconnected-cumulative}
\end{figure}

@@

In \cref{algorithm:mccis} we present a clique-inspired algorithm which finds a maximum common
connected induced subgraph isomorphism via an
association graph. If the additional
branching restrictions are removed, the core of the algorithm is what Prosser
\cite{DBLP:journals/algorithms/Prosser12} calls the ``MCSa1'' variant of a series of algorithms due
to Tomita \textit{et al.}\
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10}, using a
bitset encoding approach due to San Segundo \textit{et al.}\
\cite{DBLP:journals/cor/SegundoRJ11,DBLP:journals/ol/SegundoMRH13} (and we refer the reader to these
papers for implementation details on how to use bitsets and other data structures to implement the
colouring stage with very low constant factors).

@@

    \centering
    \input{gen-graph-connected-heatmap}
    \caption{The cumulative number of connected instances solved in under a certain time: on the
        top, 33\% undirected labelled graphs with up to 100 vertices, and in the middle, unlabelled
        and undirected graphs with up to 35 vertices. On the bottom row, an instance-by-instance
        comparison of the association and MAC both approaches, with 33\% labelled graphs on the
        left, and unlabelled and undirected graphs on the right.} \label{figure:connected-cumulative}
\end{figure}

\begin{figure}[tb]
    \centering
    \input{gen-graph-connected-cp.tex}
    \caption{A pairwise comparison of different CP techniques. The colour at a point indicates the
    number of instances with these runtimes, with darker colours representing more instances.}
        \label{figure:connected-cp}
\end{figure}

>>>>

\begin{abstract}
    The maximum common subgraph problem is to find the largest subgraph common to two given graphs.
    This problem can be solved either by constraint-based search, or by reduction to the maximum
    clique problem. We evaluate these two models using modern algorithms, and see that the best
    choice depends mainly upon whether the graphs have labelled edges. We also investigate a
    restricted branching technique for the connected version of the problem, show how to implement
    this in constraint-based and clique-inspired algorithms, and compare it to conventional
    propagation.
\end{abstract}

@@

Another widely studied approach ?? cite is to reduce the problem to finding a maximum clique (or sometimes,
equivalently, an independent set or a vertex cover ?? cite) in an association graph \cite{LeviG}; we describe
this below. Conventional wisdom is that this model is less effective in practice, but previous
experimental evaluations have used weak clique algorithms, or even maximal clique enumeration
algorithms. Maximum clique algorithms are an active research area
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10,DBLP:journals/cor/SegundoRJ11,DBLP:journals/algorithms/Prosser12,DBLP:journals/ol/SegundoMRH13,DBLP:conf/ictai/LiFX13,DBLP:journals/cor/SegundoT14,DBLP:conf/lion/SegundoLB14,DBLP:conf/cp/McCreeshP14,DBLP:journals/jco/BatsynGMP14,DBLP:journals/cor/SegundoNB15,DBLP:conf/lion/NikolaevBS15,DBLP:conf/lion/LiJX15,DBLP:journals/jcc/KoncDTRJ12,DBLP:journals/algorithms/McCreeshP13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16},
and so the first half of this paper re-evaluates the association graph model using a modern
algorithm.

@@

An interesting variant of the maximum common subgraph problem requires the selected subgraph to be
connected
\cite{DBLP:journals/tcs/Koch01,DBLP:journals/jcamd/RaymondW02a,DBLP:conf/mco/VismaraV08,Ehrlich:2011}.
Adding the connectedness requirement makes certain special cases solvable in polynomial time,
including outerplanar graphs of bounded degree \cite{DBLP:journals/algorithms/AkutsuT13} and trees
\cite{DBLP:journals/corr/DroschinskyKM16}, but the general case remains \NP-hard. The second half of
this paper looks at this variant on arbitrary graphs. In principle, for a constraint programming
approach, we may simply add a ``forms a connected graph'' global constraint
\cite{Brown:2005,DBLP:conf/cp/DoomsDD05,DBLP:conf/cp/QuesadaRD05} to the model. Alternatively, we
could use a special branching rule to maintain connectedness during search ?? cite. Neither
technique seems directly viable with an association graph encoding---however, we show that it is
possible to adapt the combined branching and bounding rule used by modern clique algorithms to
preserve connectedness during search.

Our experiments suggest that each model and technique gives broadly similar performance, although
the association graph works better when graphs have edge labels (and we discuss why), and constraint
programming models are better when edges are unlabelled. For the constraint programming model, using
both a global constraint \emph{and} the branching rule in conjunction seems to be the safest
option---although it is not always fastest, it is very rarely substantially worse than just using
one or the other, and is often substantially better.

@@

    \centering
    \input{gen-graph-unconnected-heatmap}
    \caption{The cumulative number of instances solved in under a certain time: on the top, 33\%
        labelled graphs with up to 100 vertices, and in the middle, unlabelled and undirected graphs
        with up to 35 vertices. On the bottom row, an instance-by-instance comparison of the two
        models, with 33\% labelled graphs on the left, and unlabelled and undirected graphs on
        the right.} \label{figure:unconnected-cumulative}
\end{figure}

@@

In \cref{algorithm:mccis} we present a clique-inspired algorithm which finds a maximum common
connected induced subgraph isomorphism via an
association graph. If the additional
branching restrictions are removed, the core of the algorithm is what Prosser
\cite{DBLP:journals/algorithms/Prosser12} calls the ``MCSa1'' variant of a series of algorithms due
to Tomita \textit{et al.}\
\cite{DBLP:conf/dmtcs/TomitaS03,DBLP:journals/jgo/TomitaK07,DBLP:conf/walcom/TomitaSHTW10}, using a
bitset encoding due to San Segundo \textit{et al.}\
\cite{DBLP:journals/cor/SegundoRJ11,DBLP:journals/ol/SegundoMRH13} (and we refer the reader to these
papers for implementation details on how to use bitsets and other data structures to implement the
colouring stage with very low constant factors).

@@

    \centering
    \input{gen-graph-connected-heatmap}
    \caption{The cumulative number of connected instances solved in under a certain time: on the
        top, 33\% labelled undirected graphs with up to 100 vertices, and in the middle, unlabelled
        and undirected graphs with up to 35 vertices. On the bottom row, an instance-by-instance
        comparison of the association and ``MAC both'' approaches, with 33\% labelled graphs on the
        left, and unlabelled and undirected graphs on the right.} \label{figure:connected-cumulative}
\end{figure}

\begin{figure}[p]
    \centering
    \input{gen-graph-connected-cp.tex}
    \caption{A pairwise comparison of different CP techniques. The colour at a point indicates the
    number of instances with these runtimes, with darker colours representing more instances. The
    top two rows are for 33\% labelled undirected graphs with up to 100 vertices, and the bottom two
    rows are for unlabelled and undirected graphs with up to 35 vertices.}
        \label{figure:connected-cp}
\end{figure}
